Contracts for unions and intersections are hard for Nickel

### 2021-05-25 paper.tex
Wip on flat contracts and intersection

\subsection*{Flat contracts}

@@

\subsection*{Unions}
Unions are fundamental building blocks of programs. A union holds a value whose
shape is known to be in a fixed set of alternatives: for example, a value that
can be either a boolean or a string. They correspond logically to disjunctions.
\unsure{disjunctions, right?}
Used in combination with products, they can represent arbitrary tree-like
structures, allowing to model a large class of data
(\ref{fig:union-adt-ex}).

\label{fig:union-adt-ex}
\begin{lstlisting}[title={Tree-like data structure}]
type BinaryTree =
Integer | (BinaryTree, BinaryTree)
\end{lstlisting}

@@

When programming with dynamic checks it is assumed that lack of
failure does not mean that the program works perfectly.
For instance, imagine having a function wrapped in a contract
that checks that maps positive numbers to positive numbers, like
function \nickel{f} on Figure \ref{fig:pos-to-pos}.

@@

Having a witness of size 1 for every contract failure is a nice to have property
since it means, for example, that every time that a contract is violated,
the stack trace is enough to understand how it was violated.
However, when unions get introduced, it is lost.
Consider now the program presented at Figure \ref{fig:wrong-union-function},
this program (and in particular, \nickel{f}'s contract) still presents a problem,
since \nickel{fun x => x - 7} does not map positive numbers to positive numbers,
nor does it map positive numbers to non-positive numbers, so it won't be a valid
value for the union of those contracts.

@@

Still, the program presents enough function calls to find that issue
and violate the contract,
since \nickel{f 10} is a witness of \nickel{f} failing \nickel{Positive -> NonPositive},
and \nickel{f 5} is a witness of \nickel{f} failing \nickel{Positive -> Positive}.
But, what makes this case completely different from the one presented before, is that
both function calls are needed, and, what is more, there is not a single way to
use the wrapped value \nickel{f} that would find a violation to its contract;
at least two are needed.

@@

However, this two programs behave differently, the first one will return
\nickel{(1, "False)}, without failing.
While the second one will fail, since the single application of \nickel{f 1},
applied to both \nickel{true} and \nickel{false} will be able to find the error.

Summing it up, wrapping a value with a union of higer order contracts means that
that value now has a state, that may get modified silently at every use site.
Both reviewed papers suffer from this defect since it is the semantics
of unions that interact badly with code manipulation operations,
not a particular implementation.

>>>>

\subsection*{Flat contracts}
\label{sec:flat-contracts}

@@

\subsection*{Unions}
Unions are fundamental building blocks of programs. A union holds a value whose
shape is known to be in a fixed set of alternatives: for example, a value that
can be either a boolean or a string. They correspond logically to disjunctions.
Used in combination with products, they can represent arbitrary tree-like
structures, allowing to model a large class of data
(\ref{fig:union-adt-ex}).

\change{These next few labels need to be inside figures,
they are not working}
\label{fig:union-adt-ex}
\begin{lstlisting}[title={Tree-like data structure}]
type BinaryTree =
Integer | (BinaryTree, BinaryTree)
\end{lstlisting}

@@

When programming with dynamic checks it is assumed that lack of
failure does not mean that the program works perfectly.
For instance, imagine having a function wrapped in a contract
that checks that it maps positive numbers to positive numbers, like
function \nickel{f} on Figure \ref{fig:pos-to-pos}.

@@

Having a witness of size 1 for every contract failure is a nice to have property
since it means, for example, that every time that a contract is violated,
the stack trace is enough to understand how it was violated, and to potentially
come up with a minimal example that breaks the contract.
However, when unions get introduced, it is lost.
Consider now the program presented at Figure \ref{fig:wrong-union-function},
this program (and in particular, \nickel{f}'s contract) still presents a problem,
since \nickel{fun x => x - 7} does not map positive numbers to positive numbers,
nor does it map positive numbers to non-positive numbers, so it won't be a valid
value for the union of those contracts.

@@

Still, the program presents enough function calls to find that issue
and violate the contract,
since \nickel{f 10} is a witness of \nickel{f} failing \nickel{Positive -> NonPositive},
and \nickel{f 5} is a witness of \nickel{f} failing \nickel{Positive -> Positive}.
But, what makes this case completely different from the one presented before, is that
both function calls are needed, and, what is more, there is not a way to
use the wrapped value \nickel{f} only once that would find a violation to its contract;
at least two are needed.

@@

Unexpectedly, this two programs behave differently, the first one will return
\nickel{(1, "False")}, without failing.
While the second one will fail, since the single application of \nickel{f 1},
applied to both \nickel{true} and \nickel{false} will be able to find the error.

Summing it up, wrapping a value with a union of higer order contracts means that
that value now has a state, that may get modified silently at every use site.
Both reviewed papers suffer from this defect since it is the semantics
of unions that interact badly with code manipulation operations,
not a particular implementation.
This addition of global state complicates both the work of the language
implementor, as well as the language user, since it adds non trivial
side effects to your code.
\unsure{Maybe mention CSE is actualy commonly done by hand as well}

\subsubsection*{[WIP] Flat contracts are too permissive}

Stop for a second and think how you could implement an intersection
contract. The first thought you may have is "just apply
both contracts", as shown on the Implementation of Figure \ref{fig:inter-contracts}.
Of course, if it were that simple we wouldn't be writing this paper,
this idea starts to show problems when you first combine it with
higher order contracts, since it would mean not complying with
at least one of these contracts negatively would mean breaking the
whole thing, as shown on the second example on the same Figure.

\begin{figure}[h]
\begin{lstlisting}[language=nickel, title=Implementation]
x @ A & B
(*@ $\simeq$ @*)
(x @ A) @ B
\end{lstlisting}
\begin{lstlisting}[language=nickel, title=Problem]
let g = (fun x => x) @ (Number -> Number) & (String -> String) in
g 1
\end{lstlisting}
\caption{Implementations of intersection contracts: first iteration}
\label{fig:inter-contracts}
\end{figure}

Still, not all hope is lost, we could pass some extra shared state
to each one of the sub contracts and let them only fail if the other
side of the contract has failed as well, as shown in Figure
\ref{fig:inter-contracts-2}, where the sared state is
represented by a label \nickel{l}.
This idea is still quite reasonable, for instance, when executing
\nickel{g 1}, from Figure \ref{fig:inter-contracts}, the first contract
that gets checked is the argument of \nickel{String -> String} over the function parameter, so
something like \nickel{1 @ String}.
Of course, this fails, but before raising a violation, this contract could
check if the other side of the intersection (\nickel{Number -> Number}) has
also tried to raise negative blame (on the parameter \nickel{1}), since
it hasn't, the contract would return normally, and would mark itself as failed.

\begin{figure}[h]
\begin{lstlisting}[language=nickel, title=Implementation]
x @ A & B
(*@ $\simeq$ @*)
(x @ A[l]) @ B[l]
\end{lstlisting}
\caption{Implementations of intersection contracts: second iteration}
\label{fig:inter-contracts-2}
\end{figure}

This idea, which is an oversimplification of the ideas
presented by Williams, Morris, and Wadler, seems to be spot on.
However, it presents a major problem, it's not compatible with flat contracts,
presented in Section \ref{sec:flat-contracts}.
Consider the example presented on Figure \ref{fig:inter-flat-contracts},
if the implementation presented previously were to be used, we may end up with
something like this \nickel{((fun x => x) @ (String -> String)[l]) @ C[l]}, where
\nickel{l} represents the shared state.
At this point, the intuitive thing to do is to check contract \nickel{C}, by checking
\nickel{((fun x => x) @ (String -> String)[l]) 0 == 0}.
This will fail, since applying \nickel{0} to a \nickel{String -> String} contract
fails negatively.

\begin{figure}[h]
\begin{lstlisting}[language=nickel]
let C = fun f => (f 0) == 0 in
let g = (fun x => x) @ (String -> String) & C
\end{lstlisting}
\caption{Intersection and flat contracts}
\label{fig:inter-flat-contracts}
\end{figure}

But this should not have happened, since in the program of Figure \ref{fig:inter-flat-contracts}
\nickel{g} is never applied to \nickel{0}, in fact, it is never applied.
Checking if \nickel{C} has already failed, and marking \nickel{String -> String} as
failed negatively on \nickel{l} is not enough, the contract \nickel{String -> String}
should not have been used inside the evaluation of \nickel{C}, which
adds an extra layer of complexity, contracts should be guarded, and execution
depends on the shape of the surrounding evaluation context.
